Lake Photic Zone Temperature across the Conterminous United States

### 2020-07-06 manuscript/manuscript.Rmd
Edit to MS

---
title: "Lake Photic Zone Temperature across the Conterminous United States"
author:
- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead, W. Bryan. ^*1*^,
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'

bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: inline
csl: plos.csl

abstract: Text for abstract

---

@@

During a time of unprecedented variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].  As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)[@hansen2010global], surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009) [@o2015rapid].  The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.  Ultimately, temperature changes will greatly impact every  aspects of lake resource management.  For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.   This need for accurate lake temperature forecasting will be crucial for protecting human and environmental health.

@@

##Data
We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment (NLA) (CITE).  The NLA is a stratified random sample of lakes (great than 1 ha) across the United States.  NLA sampling took place in 2007, 2012, and 2017.  This research effort used the 2007 and 2012 sample years.  The 2017 data are currently undergoing quality control before being released the public.   For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).  For each sampled lake, we used the mean temperature for all sampled depths of less than 2m (this being the depth typically considered the photic zone).

@@

Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.  Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).  Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.(ADD SENTENCE)  All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).  Model performance is reported as mean square error and adjusted R.

@@

Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.   The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean square error as a function of the number of variables.", display = "cite")`). The selected variables were the average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.

@@

The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R^2 of 0.88.  The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.  Shows precent increase in mean square error. Higher values indicate a higher impact on overall model accuracy.", display = "cite")`).  The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Partial dependence plots for selected variables.", display = "cite")`).

@@

Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States. The final model has a mean square error of 2.19 and an adjusted R^2 of 0.88. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.

>>>>

---
title: "Lake Photic Zone Temperature across the Conterminous United States"
author:
- Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
  W. Bryan. ^*1*^,
- ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
  Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
  pdf_document: default
editor_options:
  chunk_output_type: inline
csl: plos.csl
abstract: As the average global air temperature on Earth increases, surface temperatures
  of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).  The
  influence of this increased temperature touches all biotic and abiotic components
  of lentic ecosystems. Here we present a simple yet robust model of lake photic zone
  temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for
  the conterminous United States. The final model has a mean square error of 2.19
  and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average ambient air temperature,
  data obtained from the PRISM Climate Group, and longitude are the most important
  variables impacting the final model’s accuracy. Given the importance of temperature
  to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can
  be a valuable tool for researchers and lake resource managers. Daily predicted lake
  photic zone temperature for all lakes in the conterminous US can now be straightforwardly
  estimated based on basic ambient temperature and location information.
---

@@

During a time of unprecedented environmental and climatic variability, lakes can serve as sentinels and integrators in a changing world [@williamson2009lakes; @schindler2009lakes].  As the average global air temperature on Earth increases (0.15-0.20 °C per decade since 1975)[@hansen2010global], surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009) [@o2015rapid].  The influence of this increased temperature touches all biotic and abiotic components of lentic ecosystems.  Ultimately, temperature changes will greatly impact every  aspects of lake resource management.  For example, temperature, in addition to nutrients, is a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.   This need for accurate lake temperature forecasting will be crucial for protecting human and environmental health.

@@

##Data
We relied on the in situ temperature data provided in the US EPA’s National Lake Assessment (NLA) (CITE).  The NLA is a stratified random sample of lakes (great than 1 ha) across the United States.  NLA sampling took place in 2007, 2012, and 2017.  This research effort used the 2007 and 2012 sample years.  The 2017 data are currently undergoing quality control before being released the public.  Upon release, the data can easily be included to improve this model. For both sample years, we have included over 1,000 lakes across the conterminous US excluding the Great Lakes (Figure 1).  For each sampled lake, we used the mean temperature for all sampled depths of less than 2m (this being the depth typically considered the photic zone).

@@

Random forest modelling was used to not only develop a predictive model of photic zone temperature, but also used as a means of variable selection and calculate relative variable importance.  Random forest is a machine learning method that builds a consensus prediction from the assemblage of multiple tree models (here specifically 10,000 trees for the final model and 1,000 trees for the variable selection models).  Each individual tree model is constructed from a subset of the full data set and a subset of all predictor variables.All random forest modelling was conducted in R v XXXX (CITE) with the randomForest package (CITE).  Model performance is reported as mean square error and adjusted R.

@@

Using the average temperature for the upper two meters as the response variable, we initially began the variable selection process with 16 predictor variables.   The variable selection process identified a reduced model with 7 variables (`r fig_num("varsel_fig",caption="Variable selection plot for all variables. Shows percent increase in mean square error as a function of the number of variables.", display = "cite")`). The selected variables were average ambient air temperature for the sample date, sample date, longitude, average ambient air temperature for 30 day proceeding the sample date, elevation, latitude, length of lake shoreline, and the lake surface area.

@@

The final model built with the eight selected variables has a mean-squared error of 2.19 and adjusted R(n<sup>2</sup>) of 0.88.  The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.  Shows precent increase in mean square error. Higher values indicate a higher impact on overall model accuracy.", display = "cite")`).  The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Partial dependence plots for selected variables.", display = "cite")`).

@@

Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States. The final model has a mean square error of 2.19 and an adjusted R(n<sup>2</sup>) of 0.88. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be straightforwardly estimated based on basic ambient temperature and location information.
